Machine Learning for Predictive Auto-Tuning

### 2011-12-11 paper/draft1.tex
paper

The paper is organized as follows:
Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
b) empirical auto-tuning over a hill-climbing search, and
c) predictive autotuning.
%Section~\ref{sec:discussion}.
Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.

@@

Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
as a conditional optimization problem. We wish to find an implementation ($\mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
for given arguments ($\mathcal{A}$) on a particular platform ($\mathcal{C}$).
In order to perform predictive auto-tuning it is necessary to characterize these
three types of variables with {\em features}.
%In Section~\ref{sec:modeling} we see how these features are used as inputs to
%the regression model in predictive auto-tuning.

\subsection{Argument Description $\mathcal{A}$}

We describe the arguments to a filterbank correlation with the 6-tuple
$(R, C, D, F, H, W)$.
We randomly sampled arguments from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}

A library implementation of this operation should support all image and
filter sizes as well as variations due
to strided memory layouts. In such a setting would be useful to characterize
the arguments with
features such as whether the inputs are Fortran-style contiguous, C-style contiguous, or row-padded to various byte alignments.
These additional options would make our
approach of automatic auto-tuning even more important, because there would be
a greater variety in the kinds of computations and memory transfers to
perform.
Our experiments consider a somewhat simplified setting in which the arguments are always stored with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.

@@

\section{Auto-Tuning by Hill Climbing}

\subsection{Random, Grid, and Hill-Climbing Search}

@@

\begin{figure*}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over a good reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure*}

@@

\section{Boosted Regression Trees}

The regression tree is a piece-wise constant function

Boosted regression trees are

>>>>

The paper is organized as follows:
Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
b) empirical auto-tuning over a hill-climbing search, and
c) predictive autotuning.
%Section~\ref{sec:discussion}.
Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.

\section{Boosted Regression Trees}
\label{sec:brtree}

A regression tree is a piece-wise constant function from one vector space to another,
that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
We use a standard fitting procedure, which takes a set of
$(x, y)\in \mathbb{R}^k \times \mathbb{R}$ pairs and constructs a tree with a low mean squared error.
To construct each node of a regression tree,
we sort the set $\mathcal{D}$ of ($x,y$) pairs along each of the $k$ features to find the best partitioning $f_{i,\gamma}$ of the input space along feature $i$ at point $\gamma$ (Eqs.~\ref{eq:figamma}, \ref{eq:treeloss}).
\begin{align}
    f_{i,\gamma}(x) &=
    \begin{cases}
        \alpha  &\text{if $x_i < \gamma$} \\
        \beta &\text{if $x_i \geq \gamma$}
    \end{cases}
    \label{eq:figamma}\\
    i^*, \gamma^* &= \operatorname*{argmin}_{i, \gamma} \mathbb{\hat E}\left[ (y - f_{i,\gamma}(x))^2 \right]
    \label{eq:treeloss}
\end{align}
One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.

Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
In a recent empirical study of a range of machine learning regressino problems,
boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
On each boosting iteration, a regression tree is fit to the residual error remaining
after all previously-fit models have made their predictions.
There are essentially three parameters that control the boosted regression tree
training procedure:
1) the depth of tree constructed on each boosting iteration,
2) the minimum number of examples to allow at a regression tree leaf, and
3) the number of trees constructed by boosting.
We did not attempt a systematic study of the effect of these variables on performance.
We chose a maximum depth of 4,
a minimum number of examples of 10,
and 100 iterations of boosting.

@@

Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
as a conditional optimization problem in whic we seek an implementation ($b \in \mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
for given arguments ($a \in \mathcal{A}$) on a particular platform ($c \in \mathcal{C}$).
In order to perform predictive auto-tuning with a regression model,
it is necessary to characterize these
three types of variables with {\em features}.
%In Section~\ref{sec:modeling} we see how these features are used as inputs to
%the regression model in predictive auto-tuning.
We describe the arguments to a filterbank correlation with the 6-tuple
$(R, C, D, F, H, W)$.
We randomly sampled arguments (uniformly) from the following product space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}

A library implementation of this operation would ideally support all image and
filter sizes as well as variations due
to strided memory layouts. In such a setting would be useful to characterize
the arguments with
features such as whether the inputs are Fortran-style contiguous, C-style contiguous, or row-padded to various byte alignments.
These additional options would make our
approach of automatic auto-tuning even more important, because there would be
a greater variety in the kinds of computations and memory transfers to
perform.
Our experiments consider a somewhat simplified setting in which the arguments are always stored with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.

@@

For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
We did not use platform features ($c$) in our experiments, we leave the investigation
of cross-platform predictive auto-tuning for future work.

\subsection{Search Algorithms}

The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
Our implementation space includes 16000 (XXX) points, too many to search exhaustively.
\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
based on empirical autotuning on several platforms, that achieves good
performance on a variety of GPUs from older-generation cards such as the
8600GT all the way to
current high-end cards such as the GTX 580 and C2070.
We call this configuration the {\em reference} implementation.

Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
the most promising part of the configuration space. This grid iterates over all
combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.

In this work, we consider a third search algorithm based on a local search heuristic.  This {\em hill-climing} algorithm starts from the reference implementation and randomly resamples $25\%$ of the implementation parameters randomly

@@

\begin{figure}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over a good reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure}

@@
